---
title: "HP_diversity_analysis"
output: html_document
date: "2025-04-18"
---


### Load in packages

```{r, warning=FALSE}
library(tidyverse)
library("plyr")
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)

library("phyloseq")
library(rstatix)
library(cowplot)
```




### Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment. Filter out unnecessary columns

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026")) %>% 
select(sample, biosample1, experiment, passage, OTU, count, replicate, relAbundance, Phylum, Family, Genus)

e0026_clean <- e0026 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),    # remove last 4 characters
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6),
    antibiotic = if_else(str_sub(biosample1, 3, 3) == "A", 1, 0)
  )

e0026_passage8 <- e0026_clean %>% 
  filter(passage == 8)

e0026_day1 <- e0026_clean %>%  filter(str_detect(biosample1, "001") | str_detect(biosample1, "002") | str_detect(biosample1, "003") | str_detect(biosample1, "022")) %>% 
mutate(day = "001")

e0026_day29 <- e0026_clean %>%   filter(str_detect(biosample1, "029") | str_detect(biosample1, "028") | str_detect(biosample1, "027")) %>% 
mutate(day = "029")

e0026_day36 <- e0026_clean %>%  filter(str_detect(biosample1, "036")) %>% 
mutate(day = "036")

e0026_day64 <- e0026_clean %>%  filter(str_detect(biosample1, "064")) %>% 
mutate(day = "064")


# Specify output directory

OUTPUTDIR <- "./outputs"

```


### Using Phyloseq to calculate diversity statistics


```{r}

data <- readRDS("C:/abx-response-invitro/data/ps_all.rds")

hp_data_clean <- prune_taxa(taxa_sums(data) > 0, data)

e0026_obj <- subset_samples(hp_data_clean, experiment == "e0026")

```


```{r}
source("C:/abx-response-invitro/analysis/scratch/041825-computeAlphaDiversity/041825-computeAlphaDiversity.R")
source("C:/abx-response-invitro/analysis/scratch/040325-comparePrePostAbx/040325-comparePrePostAbx.R")
source("C:/abx-response-invitro/analysis/scratch/041825-computeAlphaDiversity/041825-computeAlphaDiversity.R")
source("C:/abx-response-invitro/analysis/scratch/041825-computeBetaPrePostAbx/041825-computeBetaPrePostAbx.R")
```



### Get alpha diversity metrics (ONLY RUN ONCE)

```{r}
#  # Calculate alpha diversity.
# calculateAlphaDiversity <- function(data) {
# 
#   # Use the estimate_richness function to calculate
#   # an array of alpha-diversity statistics.
#   alphaRaw <- estimate_richness(data, split=TRUE)
#   # Add the sample names.
#   alphaRaw$sample <- rownames(alphaRaw)
#   # Calculate the Shannon effective number of species.
#   alphaRaw <- alphaRaw %>%
# 	mutate(ShannonEffectiveSpecies=exp(Shannon))
#   # Tidy the dataframe.
#   alpha <- alphaRaw %>%
# 	mutate(sample=gsub("\\.","-",sample)) %>%
# 	pivot_longer(-sample, names_to="alphaStat", values_to="value")
# }
# 
# alpha_diversity <- calculateAlphaDiversity(e0026_obj)
# write_delim(alpha_diversity, paste0(OUTDIR, "alpha_diversity.txt"))

```


# Get beta diversity metrics (ONLY RUN ONCE)

```{r}
# # List the distance-calculation methods to be used.
# distMethods <- c("jsd","bray","jaccard")
# 
# # Write a function to calculate a distance matrix using the specified method
# # and convert the data into tidy format.
# calculateBeta <- function(data, distMethod) {
#   # Calculate the distance matrix using the specified method.
#   betaRaw <- distance(data, method=distMethod)
#  
#   # Convert distance matrix to a dataframe.
#   beta <- as.matrix(betaRaw)
#   beta <- as.data.frame(beta)
#   beta$sample1 <- rownames(beta)
#   # Tidy the dataframe.
#   beta <- beta %>%
# 	pivot_longer(-sample1, names_to="sample2", values_to="value")
#   beta <- beta %>%
# 	filter(sample1 != sample2) %>%
# 	mutate(method=distMethod)
# }
# 
# # Calculate the distance matrix for all of the specified methods on the species abundances.
# # Combine the distance matrices for all methods.
# betaSpecies <- do.call(rbind, lapply(distMethods, function(distMethod) {
#   print(distMethod)
#   calculateBeta(e0026_obj, distMethod)
# }))
# # Export the distance matrix generated for all of the sample pairs
# # using all of the specified methods.
# write_delim(betaSpecies, paste0(OUTDIR, "speciesBeta.txt.gz"))

```



### Define a function to extract passage information from raw sample ID

```{r}

extract_passage <- function(x) {
  ifelse(
    str_detect(x, "-mix-"),
    NA_character_,
    str_extract(x, "-[A-Z]-[08]-[A-Z0-9]+") %>%
      str_extract("[08]")
  )
}
```


### Plot change in Alpha Diversity over Time

```{r}

combined_data <- bind_rows(e0026_day1, e0026_day29, e0026_day36, e0026_day64)

# Read in alpha diversity file
alpha_diversity <- read_delim("C:/abx-response-invitro/data/alpha_diversity.txt")
data(alpha_diversity)


# Only consider samples contained within combined data
valid_samples <- combined_data %>%
  distinct(sample) %>%
  pull(sample)

# Pivot to have each statistic as its own column
alpha_diversity_wide <- alpha_diversity %>%
  filter(alphaStat %in% c("Shannon", "Chao1")) %>%
  select(sample, alphaStat, value) %>%
  pivot_wider(names_from = alphaStat, values_from = value)

# Add a column with altered sample ID
combined_data_alpha <- combined_data %>%
  distinct(sample, biosample1, day, subject, antibiotic)%>%
  mutate(xsample = paste0("X", sample))

# Left join be that sample ID
e0026_alpha <- combined_data_alpha %>%
  left_join(alpha_diversity_wide, by = c("xsample" = "sample"))

# Add day, antibiotic, and passage information
e0026_alpha <- e0026_alpha %>%
  mutate(
    day = factor(day, levels = sort(unique(day))),
    antibiotic = factor(antibiotic, levels = c(0, 1), labels = c("No", "Yes")),
    passage = extract_passage(xsample)
  )

e0026_alpha <- e0026_alpha %>% 
  filter(passage == 8)


```



```{r}
alpha_time <- ggplot(
  e0026_alpha,
  aes(x = day, y = Shannon, fill = antibiotic)
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.6,
    alpha = 0.9,
    size = 1
  )  +
  geom_jitter(
    color = "black",
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),
    shape = 21,
    stroke = 0.3,
    size = 1.5,
    alpha = 0.8
  ) +
  labs(
    title = "Change in Alpha Diversity Pre- and Post Abx",
    x = "Day",
    y = "Alpha Diversity",
    fill = "Antibiotic"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# alpha_time
# 
# ggsave("alpha_time.png", width = 14, height = 12, dpi = 300, bg = "white")

# Save the plot
windows(width = 14, height = 12)
print(alpha_time)
savePlot(file.path(OUTPUTDIR, "alpha_time.png"), type = "png", device = dev.cur())

```



### Test Significance

```{r}

# These are all the possible combinations where they vary by one value.

  valid_comparisons <- c("No_001 vs No_029", "No_001 vs No_036", "No_001 vs No_064", "No_001 vs Yes_001", "No_029 vs No_036", "No_029 vs No_064", "No_029 vs Yes_029","No_036 vs No_064", "No_036 vs Yes_036", "No_064 vs Yes_064", "Yes_001 vs Yes_029", "Yes_001 vs Yes_036", "Yes_001 vs Yes_064", "Yes_029 vs Yes_036", "Yes_029 vs Yes_064", "Yes_036 vs Yes_064")
  
# These are the ones I considered to be meaningful comparisons, but I'm actually really not sure

valid_comparisons <- c("No_001 vs No_029", "No_001 vs No_036", "No_001 vs No_064", "No_001 vs Yes_036", "No_001 vs Yes_064", "Yes_001 vs Yes_029", "Yes_001 vs Yes_036", "Yes_001 vs Yes_064")


# Step 1: Construct group labels
e0026_alpha <- e0026_alpha %>%
  mutate(abxDay = paste0(antibiotic, "_", day))

# Step 2: Ensure grouping variable is a factor
e0026_alpha$abxDay <- factor(e0026_alpha$abxDay)

# Step 3: Generate pairwise comparisons manually
comparisons <- combn(levels(e0026_alpha$abxDay), 2, simplify = FALSE)

# Step 4: Run Wilcoxon rank-sum test with rstatix
wilcox_results <- rstatix::wilcox_test(
  data = e0026_alpha,
  formula = Shannon ~ abxDay,
  comparisons = comparisons
)








# # Perform Wilcoxon rank-sum tests to investigate the significance of the
# # fold-changes in absolute abundance after antibiotic perturbation.
# # Now perform these tests for each pair of subject responses.
# statsTemp <- left_join(dataAbsAbundance %>% ungroup() %>%
#                      	mutate(subjectResponseInterval=paste0(subjectResponse, "_", intervalLong)) %>%
#                      	wilcox_test(log10FoldChangeFromPreAbx~subjectResponseInterval),
#                    	dataAbsAbundance %>% ungroup() %>%
#                      	mutate(subjectResponseInterval=paste0(subjectResponse, "_", intervalLong)) %>%
#                      	wilcox_effsize(log10FoldChangeFromPreAbx~subjectResponseInterval), by=c(".y.","group1","group2","n1","n2")) %>%
#   mutate(comparison=paste0(gsub("\n"," ",group1), " vs ", gsub("\n"," ",group2)),
#      	summary=paste0("Wilcoxon rank-sum two-sided test, ", comparison, ": n=", n1+n2,
#                     	", r=", round(effsize,2), ", p=", p, ", adjusted p-value=", p.adj, ", ", p.adj.signif,
#                     	", custom adjusted p-value=", 9*p, " (number of groups: 9)")) %>%
#   separate(group1, into=c("subjectResponse1", "interval1"), sep="_") %>%
#   separate(group2, into=c("subjectResponse2", "interval2"), sep="_") %>%
#   filter((subjectResponse1=="control" | subjectResponse2=="control") |
#        	(subjectResponse1!=subjectResponse2 & interval1==interval2))
# write(c("Compare fold-change in absolute abundance between transient and lasting responders at each time interval.",
#     	statsTemp$summary, ""), file=paste0(OUTDIR, "fig1.txt"), sep="\n", append=TRUE)

  

```




### JSD between communities of day 1&29 versus JSD of communities of day 29&36


```{r, warning=FALSE}

combined_data <- bind_rows(e0026_day1, e0026_day29, e0026_day36, e0026_day64)


beta_diversity <- read_delim("./speciesBeta.txt.gz")
data(beta_diversity)


beta_diversity_wide <- beta_diversity %>%
  select(sample1, sample2, method, value) %>%
  pivot_wider(names_from = method, values_from = value)

```



```{r}


# ensure uniqueness in combined_data
combined_data_clean <- combined_data %>%
  distinct(sample, .keep_all = TRUE)

# join sample1 metadata and check
e0026_beta_1 <- beta_diversity_wide %>%
  left_join(combined_data_clean, by = c("sample1" = "sample"))



# if the columns exist, rename them
if (all(c("day", "subject", "antibiotic") %in% names(e0026_beta_1))) {
  e0026_beta_1 <- e0026_beta_1 %>%
    dplyr::rename(
      day1 = day,
      subject1 = subject,
      antibiotic1 = antibiotic
    )
} else {
  stop("Expected columns not found after join on sample1.")
}

# Join sample2 metadata
e0026_beta <- e0026_beta_1 %>%
  left_join(combined_data_clean, by = c("sample2" = "sample"))



# Rename second set
if (all(c("day", "subject", "antibiotic") %in% names(e0026_beta))) {
  e0026_beta <- e0026_beta %>%
    dplyr::rename(
      day2 = day,
      subject2 = subject,
      antibiotic2 = antibiotic
    )
} else {
  stop("Expected columns not found after join on sample2.")
}

# Add passage extraction
e0026_beta <- e0026_beta %>%
  mutate(
    passage1 = extract_passage(sample1),
    passage2 = extract_passage(sample2)
  )



```


### Test significance


```{r}
# Filter for day combinations of interest. Maybe make a new column that has the combinations and from there decide which ones to keep.

e0026_beta_clean <- e0026_beta %>%
  mutate(day_pair = paste(day1, day2, sep = "_")) %>% 
  filter(subject1 == subject2 & day1 != day2 & passage1 == 8 & passage2 == 8)
  
# Only consider day subject A samples as day 2
e0026_beta_clean <- 
  e0026_beta_clean %>% 
  filter(antibiotic1 == 1)

# Perform statistical test by looking at significance between day pairs

left_join(e0026_beta_clean %>%
  # mutate(abxDay=paste0(antibiotic,"_",day)) %>%
  wilcox_test(jsd~day_pair),
  e0026_beta_clean %>%
  # mutate(abxDay=paste0(antibiotic,"_",day)) %>%
  wilcox_effsize(jsd~day_pair),
  by=c(".y.","group1","group2","n1","n2")) %>%
  mutate(comparison=paste0(gsub("\n"," ",group1), " vs ", gsub("\n"," ",group2)),
     	summary=paste0("Wilcoxon rank-sum two-sided test, ", comparison, ": n=", n1+n2,
                    	", r=", round(effsize,2), ", p=", p, ", adjusted p-value=", p.adj, ", ", p.adj.signif, ", custom adjusted p-value=", 4*p, " (number of groups: 4)"))

```




### Plot

```{r}

beta_time <- ggplot(
  e0026_beta_clean,
  aes(x = day_pair, y = jsd)
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.6,
    alpha = 0.9,
    size = 1
  )  +
  geom_jitter(
  color = "black",
  position = position_jitter(width = 0.1),
  shape = 21,
  stroke = 0.3,
  size = 1.5,
  alpha = 0.8
)+
  labs(
    title = "Divergence Across Samples Non-Abx Subjects",
    x = "Day Pair",
    y = "Jensen-Shannon Divergence"
  )  +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )



# Save the plot
windows(width = 14, height = 12)
print(beta_time)
savePlot(file.path(OUTPUTDIR, "beta_time.png"), type = "png", device = dev.cur())

```



### JSD between in vivo and in vitro communities (passage 0 vs. passage 8)

```{r}

# Filter for values passage numbers and day 29 samples only

jsd_p0p8 <- e0026_beta %>%
  filter(!is.na(passage1), !is.na(passage2), day1 == "029", day2 == "029")

# Perform statistical test by looking at significance between passage pairs

left_join(jsd_p0p8 %>%
  mutate(combo=paste0(passage1,"_",passage2)) %>%
  wilcox_test(jsd~combo),
  jsd_p0p8 %>%
  mutate(combo=paste0(passage1,"_",passage2)) %>%
  wilcox_effsize(jsd~combo),
  by=c(".y.","group1","group2","n1","n2")) %>%
  mutate(comparison=paste0(gsub("\n"," ",group1), " vs ", gsub("\n"," ",group2)),
     	summary=paste0("Wilcoxon rank-sum two-sided test, ", comparison, ": n=", n1+n2,
                    	", r=", round(effsize,2), ", p=", p, ", adjusted p-value=", p.adj, ", ", p.adj.signif, ", custom adjusted p-value=", 4*p, " (number of groups: 4)"))

jsd_p0p8<- jsd_p0p8 %>% 
  filter(passage1 == passage2)

jsd_p0p8_plot <- ggplot(jsd_p0p8, aes(x = factor(passage1), y = jsd, fill = factor(passage1))) +
  geom_boxplot(width = 0.6, alpha = 0.85) +
  labs(
    title = "In Vivo vs In Vitro Community JSD (Within Passage)",
    x = "Passage Pair",
    y = "Jensen-Shannon Divergence",
    fill = "Passage"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")



# Save the plot
windows(width = 14, height = 12)
print(jsd_p0p8_plot)
savePlot(file.path(OUTPUTDIR, "jsd_p0p8_plot.png"), type = "png", device = dev.cur())

```




