---
title: "scratch"
output: html_document
date: "2025-04-28"
---



# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
```


# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment. Filter out unnecessary columns
e0026 <- filter(household_data, str_detect(household_data$sample, "e0026")) %>% 
select(sample, biosample1, experiment, passage, OTU, count, replicate, relAbundance, Phylum, Family, Genus)


# Extract subject, day, household, and antibiotic information
e0026_clean <- e0026 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5), # get 
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6),
    antibiotic = if_else(str_sub(biosample1, 3, 3) == "A", 1, 0)
  )

# Divide up e0026 dataset into separate day datasets
e0026_day1 <- e0026_clean %>%  filter(str_detect(biosample1, "001") | str_detect(biosample1, "002") | str_detect(biosample1, "003") | str_detect(biosample1, "022")| str_detect(biosample1, "008")) %>% 
mutate(day = "001")

e0026_day29 <- e0026_clean %>%   filter(str_detect(biosample1, "029") | str_detect(biosample1, "028") | str_detect(biosample1, "027")) %>% 
mutate(day = "029")

e0026_day36 <- e0026_clean %>%  filter(str_detect(biosample1, "036") | str_detect(biosample1, "037")) %>% 
mutate(day = "036")

e0026_day64 <- e0026_clean %>%  filter(str_detect(biosample1, "064")| str_detect(biosample1, "063") | str_detect(biosample1, "072") | str_detect(biosample1, "059")| str_detect(biosample1, "065")) %>% 
mutate(day = "064")

# Extract the top 25 families by relative abundance to make plots better
top_families <- e0026_day29 %>%
  group_by(Family) %>%
  summarise(total_abundance = sum(relAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 25) %>%
  pull(Family)


```




### Number of Reads per Community

In the household dataset, each row contains information for a different OTU (strain). In the grouped dataset below), each row represents a sample (community) with all their OTUs and counts grouped together. 


```{r, warning=FALSE}

# Plot total distribution of reads

community_reads <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  arrange(desc(total_reads))

community_reads <- data.frame(community_reads)

ggplot(community_reads, aes(x = reorder(biosample1, -total_reads), y = total_reads))+
  geom_histogram(stat="identity")+
  labs(title = "Total read distrbution by community", x= "Community", y = "Total Reads")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())


# Find summary statistics to identify outliers 

summary(community_reads)
  
reads_first_q <- 108050
reads_third_q <- 131946

reads_first_q.data <- filter(community_reads, total_reads < 108050)

```

### Number of strains per community


```{r}

# Instead of adding up the counts, count the rows for that sample

community_strains <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(num_strains=n_distinct(OTU))

# Plot the strain number distribution across communities

ggplot(community_strains, aes(x = reorder(biosample1, -num_strains), y = num_strains))+
  geom_histogram(stat="identity")+
  labs(title = "Strain number distrbution by community", x= "Community", y = "Num Strains")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())

summary(community_strains)

strains_first_q <- 52.0  
strains_third_q <- 105.25

strains_first_q.data <- filter(community_strains, num_strains < 52)
strains_third_q.data <- filter(community_strains, num_strains > 105.25)

print(strains_first_q.data$biosample1)
print(strains_third_q.data$biosample1)

```
### XSA-036: community with the least number of strains

#### Plot number of reads per family

```{r}

XBB064_household_data <- e0026%>% 
  filter(biosample1 == "XBB-064") %>% 
  group_by(Phylum, Family) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  ggplot(aes(x = reorder(Family, -total_reads), y = log(total_reads), fill=Phylum))+
  geom_bar(stat="identity")+
  labs(title = "Family read distribution in Sample XBB-064", x= "Family Identity", y = "Log Read Count")+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

XBB064_household_data
```

### XSA-029: community with the largest number of strains

#### Plot number of reads per family

```{r}
	
	
XIB064_household_data <- e0026%>% 
  filter(biosample1 == "XIB-064") %>% 
  group_by(Phylum, Family) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  ggplot(aes(x = reorder(Family, -total_reads), y = log(total_reads), fill=Phylum))+
  geom_bar(stat="identity")+
  labs(title = "Family read distribution in Sample 	
XIB-064", x= "Family Identity", y = "Log Read Count")+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
	
XIB064_household_data

```


