---
title: "Household Project Initial Analysis"
output: html_document
date: "2025-04-03"
author: "Ana Espeleta"
---



# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
```


## Question #1: How similar are in vivo and in vitro communities?


# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment. Filter out unnecessary columns
e0026 <- filter(household_data, str_detect(household_data$sample, "e0026")) %>% 
select(sample, biosample1, experiment, passage, OTU, count, replicate, relAbundance, Phylum, Family, Genus)


# Extract subject, day, household, and antibiotic information
e0026_clean <- e0026 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6),
    antibiotic = if_else(str_sub(biosample1, 3, 3) == "A", 1, 0)
  )

# Divide up e0026 dataset into separate day datasets
e0026_day1 <- e0026_clean %>%  filter(str_detect(biosample1, "001") | str_detect(biosample1, "002") | str_detect(biosample1, "003") | str_detect(biosample1, "022")| str_detect(biosample1, "008")) %>% 
mutate(day = "001")

e0026_day29 <- e0026_clean %>%   filter(str_detect(biosample1, "029") | str_detect(biosample1, "028") | str_detect(biosample1, "027")) %>% 
mutate(day = "029")

e0026_day36 <- e0026_clean %>%  filter(str_detect(biosample1, "036") | str_detect(biosample1, "037")) %>% 
mutate(day = "036")

e0026_day64 <- e0026_clean %>%  filter(str_detect(biosample1, "064")| str_detect(biosample1, "063") | str_detect(biosample1, "072") | str_detect(biosample1, "059")| str_detect(biosample1, "065")) %>% 
mutate(day = "064")

# Extract the top 25 families by relative abundance to make plots better
top_families <- e0026_day29 %>%
  group_by(Family) %>%
  summarise(total_abundance = sum(relAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 25) %>%
  pull(Family)


# Specify output directory

OUTPUTDIR <- "./outputs"

```



### Compositional changes across passages

```{r}
# Extract only communities that have all 0-8 passages sequenced

all_passages <- household_data %>%
  filter(str_detect(household_data$sample, "e0026")) %>% 
  group_by(biosample1)  %>% 
  mutate(num_passages = n_distinct(passage)) %>% 
  filter(num_passages == 9) %>% 
  ungroup() 


my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

# Plot the total relative abundances (which are calculated per community) totaled up for each

composition_across_passages <- ggplot(all_passages, aes(x = factor(passage), y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Community Composition Across Passages" , x = "Passage Number", y = "Total Relative Abundance", fill = "Family") +
  theme_minimal()+
  theme(
  legend.position = "none"
) +
  facet_wrap(~ biosample1)


# Save the plot
windows(width = 14, height = 12)
savePlot(file.path(OUTPUTDIR, "composition_across_passages.png"), type = "png", device = dev.cur())


ggplot(all_passages, aes(x = factor(passage), y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors)


```



### Fold-change of each family in passage 0 versus passage 8 (p0/p8 for each subject in day 29)

```{r}

# Get passage 0 and passage 8 communities for day 29

p0_p8 <- e0026_day29 %>% 
  filter(passage %in% c(0, 8)) %>% 
  select(biosample1, OTU, count, passage, relAbundance, Family) %>% 
  group_by(passage, biosample1, Family) %>% 
  summarise(
    total_abundance = sum(relAbundance, na.rm = TRUE),
    .groups = "drop"
  ) %>% 
  pivot_wider(
    names_from = passage,
    values_from = total_abundance,
    names_prefix = "relAbundance_passage_"
  )%>%
  replace_na(list(relAbundance_passage_0 = 0, relAbundance_passage_8 = 0)) %>% 
  mutate(
    relAbundance_passage_0 = if_else(relAbundance_passage_0 == 0, 1e-4, relAbundance_passage_0),
    relAbundance_passage_8 = if_else(relAbundance_passage_8 == 0, 1e-4, relAbundance_passage_8)
  )

p0_p8_foldchange <- p0_p8 %>% 
  mutate(fold_change = relAbundance_passage_8/relAbundance_passage_0)

```




```{r}

# Filter the full dataset to just top 20 families
filtered_data <- p0_p8_foldchange %>%
  filter(Family %in% top_families)

fold_change_p0p8 <- filtered_data %>%
  mutate(log2_fc = log2(fold_change)) %>%
  ggplot(aes(x = fct_reorder(Family, -log2_fc), y = log2_fc, fill = Family)) +
  geom_boxplot(width = 0.9, outlier.shape = NA) +
  scale_fill_manual(values = my_colors)+
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  scale_x_discrete(expand = expansion(mult = c(0.1, 0.1))) + 
  labs(
    title = "Distribution of Log2 Fold Change per Subject (P8 vs P0)",
    x = "Subject (biosample1)",
    y = "Log2 Fold Change"
  ) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 4), legend.position = "none")

# ggsave("fold_change_p0p8.png", width = 14, height = 12, dpi = 300, bg = "white")

# Save the plot
windows(width = 14, height = 12)
print(fold_change_p0p8)
savePlot(file.path(OUTPUTDIR, "fold_change_p0p8.png"), type = "png", device = dev.cur())

```



### Summary of the most prevalent and abundant taxa across subjects at day 29.

Prevalence is the number (percentage) of samples that contain that family. Total abundance is the cumulative abundance of the different strains in a family. The mean abundance is the average between the same families across different subjects.

```{r}

# e0026_day29 <- e0026 %>%   filter(str_detect(biosample1, "029") | str_detect(biosample1, "028") | str_detect(biosample1, "027") | str_detect(biosample1, "022"))


# I want the total relative abundance of OTUs at the family-level first

summary_total_abundance <- e0026_day29 %>% 
  filter(passage == 8) %>% 
  group_by(biosample1, Family) %>% 
  summarise(
    total_abundance = sum(relAbundance, na.rm = TRUE),
    .groups = "drop"
  )

# Then, I get the mean of those total family abundances across all subjects

summary_mean_abundance <- summary_total_abundance %>% 
  group_by(Family) %>% 
  summarise(
    prevalence = n_distinct(biosample1[total_abundance > 0.001]),
    mean_abundance = mean(total_abundance, na.rm = TRUE),
    .groups = "drop"
  ) 

# Adjust prevalence to be the percent of total samples

n_samples <- e0026_day29 %>% 
  pull(biosample1) %>% n_distinct()

summary <- summary_mean_abundance %>%
  mutate(prevalence_pct = 100 * prevalence / n_samples)

summary <- summary %>%
  filter(Family %in% top_families)

  
abundance_prevalence8 <- ggplot(summary, aes(x = prevalence, y = mean_abundance, label = Family, color = Family)) +
  geom_point(aes(size = mean_abundance), alpha = 0.7) +
  geom_text_repel(size = 6, max.overlaps = Inf) +
  scale_color_manual(values = my_colors) +
  scale_x_continuous(name = "Prevalence (% of Samples)") +
  scale_y_continuous(name = "Mean Relative Abundance", trans = "log10") + 
  labs(title = "Most Prevalent and Abundant Taxa at Day 29 (Passage 8)") +
  theme_minimal()+
  theme(
  legend.position = "none"
)


# ggsave("abundance_prevalence8.png", width = 14, height = 12, dpi = 300, bg = "white")

windows(width = 14, height = 12)
print(abundance_prevalence8)
savePlot(file.path(OUTPUTDIR, "abundance_prevalence8.png"), type = "png", device = dev.cur())
```



### Comparison of community composition at days 1 and 29 (pre-abx).

How consistent is the community composition, in the absence of antibiotics?

```{r}



# Get only subject with data from both days

subjects_with_both_days <- e0026_clean %>%
  filter(day %in% c("001", "002","003", "022", "027", "028", "029")) %>%
  distinct(subject, day) %>%
  count(subject) %>%
  filter(n == 2) %>%
  pull(subject)

compare_001 <- e0026_clean %>% 
  filter(day == "001" | day == "002" | day == "003", subject %in% subjects_with_both_days) %>% 
  group_by(subject, Family) %>% 
  summarize(total_abundance = sum(relAbundance), .groups = "drop") %>%
  mutate(day = "001")

compare_029 <- e0026_clean %>% 
  filter(day == "029" | day == "028" | day == "027" | day == "022", subject %in% subjects_with_both_days) %>% 
  group_by(subject, Family) %>% 
  summarize(total_abundance = sum(relAbundance), .groups = "drop") %>%
  mutate(day = "029")

# Combine data and create two variables for each day's relative abundance

combined_data_0129 <- bind_rows(compare_001, compare_029)
wide_data <- combined_data_0129 %>%
  pivot_wider(
    names_from = day,
    values_from = total_abundance,
    names_prefix = "day_"
  ) %>%
  replace_na(list(day_001 = 0, day_029 = 0)) %>% 
  mutate(
    day_001 = if_else(day_001 == 0, 1e-4, day_001),
    day_029 = if_else(day_029 == 0, 1e-4, day_029)
  )
plot_data <- wide_data %>%
  mutate(
    log_day1 = log10(day_001),
    log_day29 = log10(day_029)
  )


day1_29_composition <- ggplot(plot_data, aes(x = log_day29, y = log_day1, color = Family)) +
  geom_point(size = 1.5, alpha = 0.9) +
  scale_color_manual(values = my_colors) +
  facet_wrap(~ subject) +
  scale_x_continuous(name = expression(log[10]*"(rel. ab.) day 29")) +
  scale_y_continuous(name = expression(log[10]*"(rel. ab.) day 1")) +
  labs(title = "Comparison of Community Composition (Relative Abundances) from Day 1 and Day 29") +
  theme_minimal() +
  theme(
    strip.text = element_text(face = "bold"),
    legend.position = "none"
  )


# day1_29_composition
# 
# ggsave("day1_29_composition.png", width = 14, height = 12, dpi = 300, bg = "white")

windows(width = 14, height = 12)
print(day1_29_composition)
savePlot(file.path(OUTPUTDIR, "day1_29_composition.png"), type = "png", device = dev.cur())


```



### Species Richness: calculate number of OTU's with a relative abundance > 0.1%

Plot the distribution of species richness at each timepoint, for abx- and non-abx subjects.

```{r}

# Combine all the day datasets

combined_data_richness <- bind_rows(e0026_day1, e0026_day29, e0026_day36, e0026_day64)

# Number of species per sample (rel abundance > 0.001)

e0026_richness <- combined_data_richness %>%
  filter(relAbundance > 0.001) %>%
  select(biosample1, OTU) %>%
  distinct() %>%
  dplyr::count(biosample1, name = "species_richness") %>%
  left_join(
    combined_data_richness %>% distinct(biosample1, day, subject, antibiotic),
    by = "biosample1"
  )


e0026_richness <- e0026_richness %>%
  mutate(
    day = factor(day, levels = sort(unique(day))),
    antibiotic = factor(antibiotic, levels = c(0, 1), labels = c("No", "Yes"))
  )


species_richness_time <- ggplot(
  e0026_richness,
  aes(x = day, y = species_richness, fill = antibiotic)
) +
  geom_boxplot(
    position = position_dodge(width = 0.75),
    width = 0.6,
    alpha = 0.9,
    size = 1
  )  +
  geom_jitter(
    color = "black",
    position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75),
    shape = 21,
    stroke = 0.3,
    size = 1.5,
    alpha = 0.8
  ) +
  labs(
    title = "Change in Species Richness Pre- and Post Abx",
    x = "Day",
    y = "Species Richness",
    fill = "Antibiotic"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

# species_richness_time
# 
# ggsave("species_richness_time.png", width = 14, height = 12, dpi = 300, bg = "white")

windows(width = 14, height = 12)
print(species_richness_time)
savePlot(file.path(OUTPUTDIR, "species_richness_time.png"), type = "png", device = dev.cur())

```


### Perform Wilcoxon rank-sum test 

Perform a Wilcoxon rank-sum test (Mann-Whitney U test) to assess whether species richness significantly differs between samples with and without antibiotics

```{r}

wilcox_results_by_day <- e0026_richness %>%
  group_by(day) %>%
  summarise(
    p_value = wilcox.test(species_richness ~ antibiotic)$p.value
  )

wilcox_results_by_day

```


## Question #2: How do communities change after antibiotics?

Plot stacked bar plots with a facet for each subject, showing compositional changes over time across timepoints, at passage 8.

### Compositional changes pre and post-abx

```{r}

# Combine day datasets again
combined_data_composition <- bind_rows(e0026_day1, e0026_day29, e0026_day36, e0026_day64)

p8_data <- combined_data_composition %>% 
  filter(passage == 8)


# Summarize relative abundance per biosample1 by Family
composition_summary <- p8_data %>%
  group_by(biosample1, Family) %>%
  summarise(relAbundance = sum(relAbundance, na.rm = TRUE), .groups = "drop") %>%
  left_join(
    p8_data %>% distinct(biosample1, day, subject, antibiotic),
    by = "biosample1"
  ) %>%
  mutate(
    day = factor(day, levels = sort(unique(day)))
  )


```



```{r}

relative_abundance_plot <- ggplot(
  composition_summary,
  aes(x = factor(day), y = relAbundance, fill = Family)
) +
  geom_bar(
    stat = "identity",
    position = "stack",
    color = "black",
    size = 0.2
  ) +
  scale_fill_manual(values = my_colors) +
  facet_wrap(~subject) +
  labs(
    title = "Relative Abundance of Microbial Families at Passage 8",
    x = "Sample",
    y = "Relative Abundance",
    fill = "Family"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

# relative_abundance_plot


# ggsave("p8_relative_abundance_plot.png", width = 14, height = 12, dpi = 300, bg = "white")

windows(width = 14, height = 12)
print(relative_abundance_plot)
savePlot(file.path(OUTPUTDIR, "relative_abundance_plot.png"), type = "png", device = dev.cur())

```


### Fold-change after abx (day36/day29) by family

Which families are the most strongly impacted by antibiotics?

```{r}

# Filter for day 29 & 36 and only antibiotic-treated subjects
pre_post_abx <- e0026_clean %>%
  filter(day %in% c("029", "036"), antibiotic == 1) %>%
  select(biosample1, subject, day, OTU, relAbundance, Family)

# Summarize relAbundance per OTU per subject per day
pre_post_abx_summary <- pre_post_abx %>%
  group_by(subject, day, OTU, Family) %>%
  summarise(total_abundance = sum(relAbundance, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(
    names_from = day,
    values_from = total_abundance,
    names_prefix = "relAbundance_day_"
  ) %>%
  replace_na(list(relAbundance_day_029 = 0, relAbundance_day_036 = 0)) %>%
  mutate(
    relAbundance_day_029 = if_else(relAbundance_day_029 == 0, 1e-4, relAbundance_day_029),
    relAbundance_day_036 = if_else(relAbundance_day_036 == 0, 1e-4, relAbundance_day_036),
    fold_change = relAbundance_day_036 / relAbundance_day_029,
    log2_fc = log2(fold_change)
  )

```



```{r}
pre_post_abx_summary <- pre_post_abx_summary %>%
  filter(Family %in% top_families)

pre_post_abx_family <- ggplot(pre_post_abx_summary, aes(x = fct_reorder(Family, -log2_fc), y = log2_fc, fill = Family)) +
  geom_boxplot(outlier.shape = NA, width = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "darkred") +
  scale_fill_manual(values = my_colors)+
  labs(
    title = "Impact of Antibiotics on Family Relative Abundance",
    x = "Family",
    y = "Log2 Fold Change (Day 36 / Day 29)"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 7),
    legend.position = "none"
  )

# pre_post_abx_family

# ggsave("pre_post_abx_family.png", width = 14, height = 12, dpi = 300, bg = "white")


windows(width = 14, height = 12)
print(pre_post_abx_family)
savePlot(file.path(OUTPUTDIR, "pre_post_abx_family.png"), type = "png", device = dev.cur())

```


