---
title: "e0029_initial_analysis"
output: html_document
date: "2025-05-06"
---


# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
```




# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment
e0029 <- filter(household_data, str_detect(household_data$sample, "e0029"))

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026"))

# Extract subject, day, household information
e0029_clean <- e0029 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6)
  )

# Specify output directory

OUTPUTDIR <- "./outputs"

# Define color palette

my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

```


### First plot composition across mixtures by biosample2 (donor community)


```{r}

# First get only the biosamples corresponding to actual mixtures
composition_across_mixes <- e0029 %>%
  # filter(!is.na(relAbundance)) %>%
  group_by(biosample2, replicate) %>%
  mutate(
    biosample1 = factor(biosample1, levels = unique(biosample1))  # sets the order of levels to match the order they appear within the current group
  ) %>%
  ungroup() # Remove the group structure to avoid it affecting during plotting



composition_across_mixes_plot <- ggplot(composition_across_mixes, aes(x = biosample1, y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Community Composition Across Mixtures",
    x = "Biosample1",
    y = "Total Relative Abundance"
  ) +
  theme_gray() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    strip.text.x = element_text(angle = 90, hjust = 0, size = 15)
  ) +
  facet_grid(replicate ~ biosample2, scales = "free_x", space = "free_x")



ggsave(
  filename = file.path(OUTPUTDIR, "composition_across_mixes.png"),
  plot = composition_across_mixes_plot,
  width = 25,
  height = 10,
  dpi = 300
)

```



### Try looking at the opposite: composition across mixtures by biosample1 (donor community)


```{r}

# First get only the biosamples corresponding to actual mixtures
composition_across_mixes <- e0029 %>%
  # filter(!is.na(relAbundance)) %>%
  group_by(biosample1, replicate) %>%  # flip group structure
  mutate(
    biosample2 = factor(biosample2, levels = unique(biosample2))  # set local order within group
  ) %>%
  ungroup()

# Why do the blank mixtures have an so many more observations?
composition_across_mixes %>%
  distinct(biosample1, biosample2, replicate) %>%  # ensure unique pairs
  group_by(biosample1) %>%
  summarise(n_biosample2 = n_distinct(biosample2)) %>%
  arrange(desc(n_biosample2))


composition_across_mixes_plot <- ggplot(composition_across_mixes, aes(x = biosample2, y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Community Composition Across Mixtures",
    x = "Biosample2",
    y = "Total Relative Abundance"
  ) +
  theme_gray() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    strip.text.x = element_text(angle = 90, hjust = 0, size = 15)
  ) +
  facet_grid(replicate ~ biosample1, scales = "free_x", space = "free_x")

ggsave(
  filename = file.path(OUTPUTDIR, "composition_across_mixes_flipped.png"),
  plot = composition_across_mixes_plot,
  width = 25,
  height = 10,
  dpi = 300
)


```




### Question #1: Was colonization resistance affected by antibiotic treatment and did it recover by day 64?

Pseudocode:

Group by mixture

- For each of these unique communities, extract all OTU identities

- For the recipient communities we can do the same by looking at biosample1+blank mixtures?

- For the donor communities we have to go back to the e0026 dataset and extract the OTUs from there

**Only use ASVs with relAbundance > 0.0001**

```{r}

# Only extract OT
e0029_colonization <- e0029 %>% 
  filter(relAbundance > 0.0001)

e0026_colonization <- e0026 %>% 
  filter(relAbundance > 0.0001)

```


**Get mixture ASVs**

```{r}


  
# Replicate 1
mixture_ASVs_rep1 <- e0029_colonization %>%
  filter(replicate == 1) %>% 
  select(mixture, OTU) %>%
  distinct() %>%                        
  group_by(mixture) %>%
  summarise(asvs = list(unique(OTU)), .groups = "drop") %>%
  deframe()

# mixture_ASVs_rep1[["XBB-029+XBA-029"]]

#Replicate 2
mixture_ASVs_rep2 <- e0029_colonization %>%
  filter(replicate == 2) %>% 
  select(mixture, OTU) %>%
  distinct() %>%                        
  group_by(mixture) %>%
  summarise(asvs = list(unique(OTU)), .groups = "drop") %>%
  deframe()

```


**Get recipient ASVs**

```{r}



# Replicate 1
recipient_ASVs_rep1 <- e0029_colonization %>% 
  filter(replicate == 1, biosample2 == "blank", biosample1 != "blank") %>% 
  select(biosample1, OTU) %>%
  distinct() %>%                        
  group_by(biosample1) %>%
  summarise(asvs = list(unique(OTU)), .groups = "drop") %>%
  deframe()

# Replicate 2
recipient_ASVs_rep2 <- e0029_colonization %>% 
  filter(replicate == 2, biosample2 == "blank", biosample1 != "blank") %>% 
  select(biosample1, OTU) %>%
  distinct() %>%                        
  group_by(biosample1) %>%
  summarise(asvs = list(unique(OTU)), .groups = "drop") %>%
  deframe()

# recipient_ASVs_rep1[["XBA-029"]]

```


**Get donor ASVs**

```{r}

# Extract only the donor communities
donor_communities <- e0029_colonization %>%
  filter(!is.na(biosample2), biosample2 != "blank") %>%
  pull(biosample2) %>%
  unique()

# Replicate 1
donor_ASVs_rep1 <- e0026_colonization %>% 
  filter(biosample1 %in% donor_communities) %>% 
  select(biosample1, OTU) %>%
  distinct() %>%
  group_by(biosample1) %>%
  summarise(asvs = list(unique(OTU)), .groups = "drop") %>%
  deframe()


```



### Define a function to categorize colonization

this is how you get an overall intersection: global_intersection <- intersect(
  unlist(list1),
  unlist(list2)
)


```{r}

for (mix in names(mixture_ASVs_rep1)) {
  otus <- mixture_ASVs_rep1[[mix]]
  cat("Mixture:", mix, "\n")
  print(otus)
}



```



