---
title: "Household Project Initial Analysis"
output: html_document
date: "2025-04-03"
author: "Ana Espeleta"
---



# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
```


# Load in the Household Project dataset

```{r}
household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment. Filter out unnecessary columns

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026")) %>% 
select(sample, biosample1, experiment, OTU, count, replicate, relAbundance)


e0029 <- filter(household_data, str_detect(household_data$sample, "e0029")) %>% 
select(sample, biosample1, biosample2, experiment, OTU, count, replicate, relAbundance)


```


## Experiment 0026 Summary Statistics


### Number of Reads per Community

In the household dataset, each row contains information for a different OTU (strain). In the grouped dataset below), each row represents a sample (community) with all their OTUs and counts grouped together. 

Q: Do we want to instead count how many OTUs each community has by counting the rows instead not summing it all up?

```{r, warning=FALSE}

# Plot total distribution of reads

community_reads <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  arrange(desc(total_reads))

community_reads <- data.frame(community_reads)

ggplot(community_reads, aes(x = reorder(biosample1, -total_reads), y = total_reads))+
  geom_histogram(stat="identity")+
  labs(title = "Total read distrbution by community", x= "Community", y = "Total Reads")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())


# Find summary statistics to identify outliers 

summary(community_reads)
  
reads_first_q <- 108050
reads_third_q <- 131946

reads_first_q.data <- filter(community_reads, total_reads < 108050)

```

### Number of strains per community


```{r}

# Instead of adding up the counts, count the rows for that sample

community_strains <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(num_strains=n())

# Plot the strain number distribution across communities

ggplot(community_strains, aes(x = reorder(biosample1, -num_strains), y = num_strains))+
  geom_histogram(stat="identity")+
  labs(title = "Strain number distrbution by community", x= "Community", y = "Num Strains")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())

summary(community_strains)

strains_first_q <- 169.0  
strains_third_q <- 307.0

strains_first_q.data <- filter(community_strains, num_strains < 169)
strains_third_q.data <- filter(community_strains, num_strains > 307)

print(strains_first_q.data$biosample1)
print(strains_third_q.data$biosample1)

# Check to make sure everything was computed right and there is no overlap between the two groups

length(intersect(strains_first_q.data, strains_third_q.data)) > 0
```
### Looking at the community with the least amount of strains (XSA-036). Plotting the counts per strain

```{r}

XSA036_household_data <- household_data%>% 
  filter(biosample1 == "XSA-036")

ggplot(XSA036_household_data, aes(x = reorder(biosample1, -total_reads), y = total_reads))+
  geom_histogram(stat="identity")+
  labs(title = "Total read distrbution by community", x= "Community", y = "Total Reads")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())
```



### Species Richness

```{r}



```

