---
title: "Household Project Initial Analysis"
output: html_document
date: "2025-04-03"
author: "Ana Espeleta"
---



# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
```


# Load in the Household Project dataset

```{r}
household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment. Filter out unnecessary columns

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026")) %>% 
  filter(passage == 8) %>% 
select(sample, biosample1, experiment, passage, OTU, count, replicate, relAbundance, Family, Phylum)


e0029 <- filter(household_data, str_detect(household_data$sample, "e0029")) %>% 
select(sample, biosample1, biosample2, experiment, OTU, count, replicate, relAbundance, Family, Phylum)


day2_data <- filter(household_data, str_detect(household_data$biosample1, "002"))
day29_data <- filter(household_data, str_detect(household_data$biosample1, "029"))
day36_data <- filter(household_data, str_detect(household_data$biosample1, "036"))
day64_data <- filter(household_data, str_detect(household_data$biosample1, "064"))

```


## Experiment 0026 Summary Statistics


### Number of Reads per Community

In the household dataset, each row contains information for a different OTU (strain). In the grouped dataset below), each row represents a sample (community) with all their OTUs and counts grouped together. 


```{r, warning=FALSE}

# Plot total distribution of reads

community_reads <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  arrange(desc(total_reads))

community_reads <- data.frame(community_reads)

ggplot(community_reads, aes(x = reorder(biosample1, -total_reads), y = total_reads))+
  geom_histogram(stat="identity")+
  labs(title = "Total read distrbution by community", x= "Community", y = "Total Reads")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())


# Find summary statistics to identify outliers 

summary(community_reads)
  
reads_first_q <- 108050
reads_third_q <- 131946

reads_first_q.data <- filter(community_reads, total_reads < 108050)

```

### Number of strains per community


```{r}

# Instead of adding up the counts, count the rows for that sample

community_strains <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(num_strains=n_distinct(OTU))

# Plot the strain number distribution across communities

ggplot(community_strains, aes(x = reorder(biosample1, -num_strains), y = num_strains))+
  geom_histogram(stat="identity")+
  labs(title = "Strain number distrbution by community", x= "Community", y = "Num Strains")+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank())

summary(community_strains)

strains_first_q <- 52.0  
strains_third_q <- 105.25

strains_first_q.data <- filter(community_strains, num_strains < 52)
strains_third_q.data <- filter(community_strains, num_strains > 105.25)

print(strains_first_q.data$biosample1)
print(strains_third_q.data$biosample1)

```
### XSA-036: community with the least number of strains

#### Plot number of reads per family

```{r}

XBB064_household_data <- e0026%>% 
  filter(biosample1 == "XBB-064") %>% 
  group_by(Phylum, Family) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  ggplot(aes(x = reorder(Family, -total_reads), y = log(total_reads), fill=Phylum))+
  geom_bar(stat="identity")+
  labs(title = "Family read distribution in Sample XBB-064", x= "Family Identity", y = "Log Read Count")+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

XBB064_household_data
```

### XSA-029: community with the largest number of strains

#### Plot number of reads per family

```{r}
	
	
XIB064_household_data <- e0026%>% 
  filter(biosample1 == "XIB-064") %>% 
  group_by(Phylum, Family) %>% 
  summarise(total_reads = sum(as.numeric(count))) %>% 
  ggplot(aes(x = reorder(Family, -total_reads), y = log(total_reads), fill=Phylum))+
  geom_bar(stat="identity")+
  labs(title = "Family read distribution in Sample 	
XIB-064", x= "Family Identity", y = "Log Read Count")+
  theme_light()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
	
XIB064_household_data

```



### Species Richness: calculate number of OTU's with a relative abundance > 0.1%

```{r}

# apply condition to see which rows are > 0.1%, count the true conditions. For each community plot how much it changes over time after antibiotics

#Plot the distribution of species richness at each time point for samples with and without antibiotic (facet by subject)
 
e0026_richness <- e0026 %>% 
  group_by(biosample1) %>% 
  summarise(richness = sum(relAbundance > 0.001))	



```



### Compositional changes across passages

```{r}
# Extract only communities that have all 0-8 passages sequenced

all_passages <- household_data %>%
  filter(str_detect(household_data$sample, "e0026")) %>% 
  group_by(biosample1)  %>% 
  mutate(num_passages = n_distinct(passage)) %>% 
  filter(num_passages == 9) %>% 
  ungroup() 


my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

# Plot the total relative abundances (which are calculated per community) totaled up for each

ggplot(all_passages, aes(x = factor(passage), y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(title = "Community composition across passages" , x = "Passage Number", y = "Total Relative Abundance", fill = "Family") +
  theme_minimal()+
  theme(
  legend.position = "none"
) +
  facet_wrap(~ biosample1)

ggplot(all_passages, aes(x = factor(passage), y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors)

# I think the reason that the bars are 12 units high is because there are 12 communities being represented, which each of their relative abundances adding up to 1


```


