---
title: "e0029_initial_analysis"
output: html_document
date: "2025-05-06"
---


# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
library(purrr)
library(patchwork)
library(foreach)
library(forcats)
library(cowplot)
```




# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment
e0029 <- filter(household_data, str_detect(household_data$sample, "e0029"))

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026"))

# Extract subject, day, household information
e0029_clean <- e0029 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6)
  )

# Specify output directory

OUTPUTDIR <- "./outputs"

# Define color palette

my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

```


### First plot composition across mixtures by biosample2 (donor community)


```{r}

# First get only the biosamples corresponding to actual mixtures
composition_across_mixes <- e0029 %>%
  filter(
    !(biosample2 == "blank") |
    (biosample2 == "blank" & !well %in% c("G11", "H11"))) %>%
  filter(
    biosample1 != "blank" | 
    (plate == "e0029.B.5" & well %in% c("G12", "H12"))
  ) %>% 
  group_by(biosample2, replicate) %>%
  mutate(
    biosample1 = factor(biosample1, levels = unique(biosample1))
  ) %>%
  ungroup()


composition_across_mixes_plot <- ggplot(composition_across_mixes, aes(x = biosample1, y = relAbundance, fill = Family)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Community Composition Across Mixtures",
    x = "Biosample1",
    y = "Total Relative Abundance"
  ) +
  theme_gray() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10),
    strip.text.x = element_text(angle = 90, hjust = 0, size = 15)
  ) +
  facet_grid(replicate ~ biosample2, scales = "free_x", space = "free_x")



ggsave(
  filename = file.path(OUTPUTDIR, "composition_across_mixes.png"),
  plot = composition_across_mixes_plot,
  width = 25,
  height = 10,
  dpi = 600
)

```



### Try looking at the opposite: composition across mixtures by biosample1 (donor community)


```{r}

# First get only the biosamples corresponding to actual mixtures. For blanks, only get the first well of the duplicates
composition_across_mixes <- e0029 %>%
  filter(
    !(biosample2 == "blank") |
    (biosample2 == "blank" & !well %in% c("G11", "H11"))) %>%
  filter(
    biosample1 != "blank" | 
    (plate == "e0029.B.5" & well %in% c("G12", "H12"))
  ) %>% 
  group_by(biosample1, replicate) %>%
  mutate(
    biosample2 = factor(biosample2, levels = unique(biosample2))
  ) %>%
  ungroup()



# Why do the blank mixtures have an so many more observations?
composition_across_mixes %>%
  distinct(biosample1, biosample2, replicate) %>%  # ensure unique pairs
  group_by(biosample1) %>%
  summarise(n_biosample2 = n_distinct(biosample2)) %>%
  arrange(desc(n_biosample2))



composition_across_mixes_plot <- ggplot(
  composition_across_mixes,
  aes(x = forcats::fct_relevel(factor(biosample2), "blank"), y = relAbundance, fill = Family)
) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Community Composition Across Mixtures",
    x = "Biosample2",
    y = "Total Relative Abundance"
  ) +
  theme_gray() + 
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 10, face = "bold"),
    strip.text.x = element_text(angle = 90, hjust = 0, size = 10)
  ) +
  facet_grid(replicate ~ biosample1, scales = "free_x", space = "free_x")



ggsave(
  filename = file.path(OUTPUTDIR, "composition_across_mixes_flipped.png"),
  plot = composition_across_mixes_plot,
  width = 25,
  height = 10,
  dpi =600
)

```




### Question #1: Was colonization resistance affected by antibiotic treatment and did it recover by day 64?

Pseudocode:

Group by mixture

- For each of these unique communities, extract all OTU identities

- For the recipient communities we can do the same by looking at biosample1+blank mixtures?

- For the donor communities we have to go back to the e0026 dataset and extract the OTUs from there

- Recalculate relative abundances for datasets (since there are replicates) and b-mixes and filter for those asvs >0.0001

**Only use ASVs with relAbundance > 0.001**

```{r}

# Have filter for relative abundance in actual calculations to avoid losing information
# Only extract OTU

limit_of_detection <- 0.001

```


**Create a mixture ASVs, recipient ASVs, and donor ASVs dataset**

```{r}

# Get mixture communities
mixture_ASVs <- e0029 %>% 
  filter(relAbundance > limit_of_detection)


# Get recipient communities
recipient_ASVs <- e0029 %>% 
  filter(relAbundance > limit_of_detection) %>%   
  filter(biosample2 == "blank") %>% 
  filter(
    !(biosample2 == "blank") |
    (biosample2 == "blank" & !well %in% c("G11", "H11"))) %>%
  filter(
    biosample1 != "blank" | 
    (plate == "e0029.B.5" & well %in% c("G12", "H12"))
  )


# Extract donor community IDs from e0029
donor_communities <- e0029 %>%
  pull(biosample2) %>%
  unique()

donor_communities

# Extract single donor OTUs from e0026
single_donor_ASVs <- e0026 %>%
  filter(biosample1 %in% donor_communities, passage == "8")
  
# We'll deal with the b-mix and two-donor communities later
donor_combinations <- list(
  c("XBB-029", "XKB-029"),
  c("XCB-029", "XBB-029"),
  c("XDB-029", "XCB-029"),
  c("XEB-029", "XDB-029"),
  c("XFB-029", "XEB-029"),
  c("XGB-029", "XFB-029"),
  c("XHB-029", "XGB-029"),
  c("XIB-029", "XHB-029"),
  c("XJB-029", "XIB-029"),
  c("XKB-029", "XJB-029") 
)


```


**Get two-donor communities**

```{r}
# Loop and collect into one dataframe. map_dfr binds all the donor pair dataframes together at the end. Iterates over each element in the list donor_combinations (i.e., each donor pair). Applies the anonymous "function(donor_pair)" to each element. Combines all returned dataframes by row (dfr = data frame row-binding)

two_donor_ASVs <- map_dfr(donor_combinations, function(donor_pair) {
  pair_name <- paste(donor_pair, collapse = "+")
  
  two_donor_ASVs <- single_donor_ASVs %>%
    filter(biosample1 %in% donor_pair)
  
  combined_otus <- two_donor_ASVs %>%
    group_by(OTU) %>%
    summarize(total_abundance = sum(relAbundance), .groups = "drop") %>%
    mutate(
      avg_abundance = total_abundance / 2, # avg_abundance ends up being the abundance of each OTU
      relAbundance_final = avg_abundance / sum(avg_abundance) # final abundance gets the OTU abundance relative to all other OTUs in the mixture
    ) %>%
    select(OTU, relAbundance_final) %>%
    filter(relAbundance_final > limit_of_detection) %>%
    mutate(mixture = pair_name)  # Add mixture ID
  
  return(combined_otus) # donor pair data frame
})

# within each mixture community, the relative abundances should this sum to one
two_donor_ASVs %>% 
  group_by(mixture) %>% 
  summarise(total = sum(relAbundance_final))



```


**Get b-mix communities**


```{r}

two_donor_ASVs

single_donor_ASVs
```


**Join together all donor community dataframes**

```{r}


```



Write a function to take in a specific mixture and output whether there was colonization or not for each OTU in the mixture (THIS ONE WORKS!)

**Testing on a specific case**
```{r}

# function takes the id and replicate number, filters the community datasets for those ids and finds intersection between the otu vectors to see if they are shared across communities 

get_colonization <- function(mix_id, donor_id, recipient_id, replicate) {
  
  # Subset mixture and recipient ASVs for the given replicate
  mix_asvs_subset <- mixture_ASVs %>% 
    filter(mixture == mix_id, replicate == replicate)
  
  recipient_asvs_subset <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == replicate)
  
  donor_asvs_subset <- single_donor_ASVs %>%  # change this when you have donor_asvs with mixed donors
    filter(biosample1 == donor_id)
  
  # Get shared OTUs between donor and mixture.  Add colonization column: 1 if OTU is shared with donor, 0 otherwise
  mix_asvs_colonization <- mix_asvs_subset %>% 
    mutate(colonization = ifelse(OTU %in% donor_asvs_subset$OTU & !(OTU %in% recipient_asvs_subset$OTU), 1, 0))

  return(mix_asvs_colonization)
}


XBB_XEA_colonization <- get_colonization(
  mix_id = "XBB-029+XEA-029",
  donor_id = "XBB-029",
  recipient_id = "XEA-029",
  replicate = 1
)

```




**Loop through all mixtures to get final e0029 colonization dataset**

```{r}

# use foreach() and rbind() to combine the subset mix_asvs_colonization datasets

# Filter just the mixtures you want to process
mixture_ids <- unique(mixture_ASVs %>%
  filter(biosample2 %in% single_donor_ASVs$biosample1) %>%
  pull(mixture))

# Apply get_colonization() across all mixture IDs
colonization_results <- foreach(mix_id = mixture_ids, .combine = bind_rows) %do% {
  ids <- unlist(strsplit(mix_id, "\\+"))
  
  donor_id <- ids[1]
  recipient_id <- ids[2]
  
  # Run your function (assumes replicate 1 here)
  result <- get_colonization(
    mix_id = mix_id,
    donor_id = donor_id,
    recipient_id = recipient_id,
    replicate = 1
  )
  
  # Add metadata to the result
  result <- result %>%
    mutate(mixture = mix_id)
  
  return(result)
}


colonization_results %>% 
  group_by(mixture) %>% 
  summarize(total_colonizers = sum(colonization))


```


### Plots!!! (FOR NOW ONLY ON REPLICATE 1)


**Looking at specific cases and verifying colonization algo**

**Day 029**

```{r}
theme_set(theme_cowplot())

mixture_ASVs_XBBXBA <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-029", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Mixture",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


mixture_ASVs_XBBXBA_colonized <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-029", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(colonization))) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
  labs(
    title = "Colonized",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


recipient_ASVs_XBBXBA <- recipient_ASVs %>% 
  filter(biosample1 == "XBA-029", replicate == 1)%>% 
  ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Recipient",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))

donor_ASVs_XBBXBA <- single_donor_ASVs %>% 
  filter(biosample1 == "XBB-029")%>%
  ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Donor",
    x = "",
    y = "Total Relative Abundance"
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


row_of_plots <- plot_grid(
  donor_ASVs_XBBXBA,
  recipient_ASVs_XBBXBA,
  mixture_ASVs_XBBXBA,
  mixture_ASVs_XBBXBA_colonized,
  ncol = 4
)

# Add a centered title above the row
preantibiotic_plot <- plot_grid(
  ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
  row_of_plots,
  ncol = 1,
  rel_heights = c(0.1, 1)  # Reserve top space for title
)

# Display the result
preantibiotic_plot




```

**Day 036**

```{r}

theme_set(theme_cowplot())
mixture_ASVs_XBBXBA36 <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-036", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Mixture",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))




mixture_ASVs_XBBXBA36_colonized <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-036", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(colonization))) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
  labs(
    title = "Colonized",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


recipient_ASVs_XBBXBA36 <- recipient_ASVs %>% 
  filter(biosample1 == "XBA-036", replicate == 1)%>% 
  ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Recipient",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank())

donor_ASVs_XBBXBA <- single_donor_ASVs %>% 
  filter(biosample1 == "XBB-029")%>%
  ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Donor",
    x = "",
    y = "Total Relative Abundance"
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


row_of_post_plots <- plot_grid(
  donor_ASVs_XBBXBA,
  recipient_ASVs_XBBXBA36,
  mixture_ASVs_XBBXBA36,
  mixture_ASVs_XBBXBA36_colonized,
  ncol = 4
)

# Add a centered title above the row
postantibiotic_plot <- plot_grid(
  ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
  row_of_post_plots,
  ncol = 1,
  rel_heights = c(0.1, 1)
)

# Display
postantibiotic_plot

# MAKE THIS LIGHT!
ggsave(
  filename = file.path(OUTPUTDIR, "postantibiotic_compositionsXBBXBA.png"),
  plot = postantibiotic_plot,
  width = 20,
  height = 10,
  dpi =600,
)



```

**Day 064**

```{r}

theme_set(theme_cowplot())
mixture_ASVs_XBBXBA64 <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-064", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Mixture",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


mixture_ASVs_XBBXBA64_colonized <- colonization_results %>% 
  filter(mixture == "XBB-029+XBA-064", replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(colonization))) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
  labs(
    title = "Colonized",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


recipient_ASVs_XBBXBA64 <- recipient_ASVs %>% 
  filter(biosample1 == "XBA-064", replicate == 1)%>% 
  ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Recipient",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


row_of_post64_plots <- plot_grid(
  donor_ASVs_XBBXBA,
  recipient_ASVs_XBBXBA64,
  mixture_ASVs_XBBXBA64,
  mixture_ASVs_XBBXBA64_colonized,
  ncol = 4
)

# Add a centered title above the row
postantibiotic64_plot <- plot_grid(
  ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
  row_of_post64_plots,
  ncol = 1,
  rel_heights = c(0.1, 1)
)

# Display
postantibiotic64_plot

```


**Join plots with patchwork**

```{r}

library(patchwork)

preantibiotic_plot <- preantibiotic_plot + theme_cowplot()
postantibiotic_plot <- postantibiotic_plot + theme_cowplot()
postantibiotic64_plot <- postantibiotic64_plot + theme_cowplot()


combined_plot <- preantibiotic_plot /
                 postantibiotic_plot /
                 postantibiotic64_plot

ggsave(
  filename = file.path(OUTPUTDIR, "compositions_XBBXBA.png"),
  plot = combined_plot,
  width = 10,
  height = 10,
  dpi = 600,
  bg = "white"
)

```



**Create function extract specific mixture and display its composition**

```{r}

theme_set(theme_cowplot())

subject_composition <- function(mix_id, donor_id, recipient_id, replicate) {
  
  # Subset mixture and recipient ASVs for the given replicate
  mix_asvs_subset <- mixture_ASVs %>% 
    filter(mixture == mix_id, replicate == replicate)
  
  recipient_asvs_subset <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == replicate)
  
  mixture_ASVs_mix <- colonization_results %>% 
  filter(mixture == mix_id, replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Mixture",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


  mixture_ASVs_colonized <- colonization_results %>% 
    filter(mixture == mix_id, replicate == 1) %>% 
    ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(colonization))) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
    labs(
      title = "Colonized",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))

  
  recipient_ASVs <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == 1)%>% 
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Recipient",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))

  donor_ASVs <- single_donor_ASVs %>% 
    filter(biosample1 == donor_id)%>%
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Donor",
      x = "",
      y = "Total Relative Abundance"
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))


  row_of_plots <- plot_grid(
    donor_ASVs,
    recipient_ASVs,
    mixture_ASVs_mix,
    mixture_ASVs_colonized,
    ncol = 4
  )
  
  # Add a centered title above the row
  final_plot <- plot_grid(
    ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
    row_of_plots,
    ncol = 1,
    rel_heights = c(0.1, 1)  # Reserve top space for title
  )

  return(final_plot)
}


```



**Function call**

```{r}

subject_composition_plot_day29 <- subject_composition("XKB-029+XEA-029", "XKB-029", "XEA-029", 1) + theme_cowplot()
subject_composition_plot_day36 <- subject_composition("XKB-029+XEA-036", "XKB-029", "XEA-036", 1) + theme_cowplot()
subject_composition_plot_day64 <- subject_composition("XKB-029+XEA-064", "XKB-029", "XEA-064", 1) + theme_cowplot()

combined_plot <- subject_composition_plot_day29 /
                 subject_composition_plot_day36 /
                 subject_composition_plot_day64

ggsave(
  filename = file.path(OUTPUTDIR, "compositions_XKBXEA.png"),
  plot = combined_plot,
  width = 10,
  height = 10,
  dpi = 600,
  bg = "white"
)


```


**Get species richness for each community**

```{r}

theme_set(theme_cowplot())

# Create recipient richness dataset
recipient_richness <- recipient_ASVs %>%
  filter(biosample1 != "blank") %>% 
  select(biosample1, OTU) %>%
  distinct() %>%
  dplyr::count(biosample1, name = "species_richness")  %>% # count the distinct OTUs
  mutate(community = "recipient", 
         day = as.integer(str_extract(biosample1, "\\d{2}$"))) %>%
  rename(sample = biosample1)



# Create mixture richness dataset
mixture_richness <- mixture_ASVs %>%
  filter(biosample2 %in% single_donor_ASVs$biosample1) %>%
  select(mixture, OTU) %>%
  distinct() %>%
  dplyr::count(mixture, name = "species_richness") %>% #count the distinct OTUs
  mutate(community = "mixture", 
         day = as.integer(str_extract(mixture, "\\d{2}$"))) %>%
  rename(sample = mixture)


# Join richness datasets together to plot
combined_richness <- bind_rows(recipient_richness, mixture_richness)


# Plot
richness_day_plot <- 
  ggplot(combined_richness, aes(x = factor(day), y = species_richness, fill = community)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8, color = "black", width = 0.6) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.6) +
  labs(
    title = "Change in Species Richness Pre- and Post-Abx",
    x = "Day",
    y = "Species Richness",
    fill = "Community Type"
  )


richness_day_plot

ggsave(
  filename = file.path(OUTPUTDIR, "richness_by_day.png"),
  plot = richness_day_plot,
  width = 10,
  height = 5.5,
  dpi = 600,
  bg = "white"
)

```


### How does the recipient's species richness compare to the number of colonizers in the mixture?


```{r}


# we need the colonization_results and the recipient_richness datasets. From there, we have to join them by biosample1 

total_colonization <- colonization_results %>% 
  group_by(mixture) %>% 
  summarize(total_colonizers = sum(colonization)) %>% 
  mutate(recipient = str_sub(mixture, -7, -1))

combined_richness <- left_join(total_colonization, recipient_richness,
  by = c("recipient" = "sample") # match recipient to sample
) %>% 
  select(-community)  %>% 
  mutate(recipient = str_sub(recipient, -7, -5))
  


colonization_richness_plot <- combined_richness %>% 
  ggplot(aes(x = species_richness, y = total_colonizers, color = factor(day))) + 
  geom_point() +
  geom_text_repel(aes(label = recipient), size = 2)+
  labs(
    x = "Species Richness of Recipient",
    y = "Number of Colonizers in Mixture"
  )


ggsave(
  filename = file.path(OUTPUTDIR, "colonization_vs_richness.png"),
  plot = colonization_richness_plot,
  width = 9,
  height = 6,
  dpi = 600,
  bg = "white"
)


# Could I instead use color for the subject and shape of dot for day?

```


**Boxplots of total colonizers by mixture and day**


```{r}

total_colonization <- colonization_results %>% 
  group_by(mixture, biosample1) %>% 
  summarize(total_colonizers = sum(colonization)) %>% 
  mutate(day = str_sub(mixture, -2), recipient = str_sub(biosample1, 0, 3))


# Plot
colonization_day_plot2 <- 
  ggplot(total_colonization, aes(x = factor(day), y = total_colonizers, fill = factor(day))) +
  geom_boxplot() +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.6) +
  labs(
    title = "Number of Colonizers per Mixture by Day",
    x = "Day",
    y = "Number of colonizers",
    fill = ""
  )+
  facet_wrap(~recipient)

colonization_day_plot2

ggsave(
  filename = file.path(OUTPUTDIR, "colonization_by_day.png"),
  plot = colonization_day_plot2,
  width = 9,
  height = 6,
  dpi = 600,
  bg = "white"
)


```





