---
title: "e0029 Divergence and Colonization Analysis"
output: html_document
date: "2025-05-21"
---


# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
library(purrr)
library(patchwork)
library(foreach)
library(forcats)
library(cowplot)
```


# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment
e0029 <- filter(household_data, str_detect(household_data$sample, "e0029"))

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026"))

# Extract subject, day, household information
e0029_clean <- e0029 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6)
  )

# Specify output directory

OUTPUTDIR <- "./e0029_outputs"

# Define color palette

my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

```






**Computing colonization as an efficacy metric**

```{r}

get_colonization <- function(mix_id, donor_id, recipient_id, replicate) {
  mix_asvs_subset <- mixture_ASVs %>% 
    filter(mixture == mix_id, replicate == 1)
  
  recipient_asvs_subset <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == 1)
  
  donor_asvs_subset <- single_donor_ASVs %>%  
    filter(biosample1 == donor_id)
  
  # potential colonizers are those present in the donor but not in the recipient
  potential_colonizers <- donor_asvs_subset$OTU[!(donor_asvs_subset$OTU %in% recipient_asvs_subset$OTU)]
  
  donor_asvs_potential <- donor_asvs_subset %>% 
    filter(OTU %in% potential_colonizers)

  # actual colonizers are the subset of potentials that did make it into the mix
  mix_asvs_colonization <- mix_asvs_subset %>% 
    mutate(actual_colonizer = ifelse(OTU %in% potential_colonizers, 1, 0))
  
  # return both the dataframe and potential colonizer list
  return(list(
    colonization_df = mix_asvs_colonization,
    donor_asvs_potential = donor_asvs_potential
  ))
}

```




**Create a mixture ASVs, recipient ASVs, and donor ASVs dataset**

```{r}

limit_of_detection <- 0.001

# Get mixture communities
mixture_ASVs <- e0029 %>% 
  filter(relAbundance > limit_of_detection)


# Get recipient communities
recipient_ASVs <- e0029 %>% 
  filter(relAbundance > limit_of_detection) %>%   
  filter(biosample2 == "blank") %>% 
  filter(
    !(biosample2 == "blank") |
    (biosample2 == "blank" & !well %in% c("G11", "H11"))) %>%
  filter(
    biosample1 != "blank" | 
    (plate == "e0029.B.5" & well %in% c("G12", "H12"))
  )


# Extract donor community IDs from e0029
donor_communities <- e0029 %>%
  pull(biosample2) %>%
  unique()

donor_communities

# Extract single donor OTUs from e0026
single_donor_ASVs <- e0026 %>%
  filter(biosample1 %in% donor_communities, passage == "8")
  
# We'll deal with the b-mix and two-donor communities later
donor_combinations <- list(
  c("XBB-029", "XKB-029"),
  c("XCB-029", "XBB-029"),
  c("XDB-029", "XCB-029"),
  c("XEB-029", "XDB-029"),
  c("XFB-029", "XEB-029"),
  c("XGB-029", "XFB-029"),
  c("XHB-029", "XGB-029"),
  c("XIB-029", "XHB-029"),
  c("XJB-029", "XIB-029"),
  c("XKB-029", "XJB-029") 
)


```


**Get a final dataset with actual colonizers**

```{r}

# initialize results table
actual_colonizers_results <- tibble()
potential_colonizers_results <- tibble()

# Filter just the mixtures you want to process
mixture_ids <- unique(mixture_ASVs %>%
  filter(biosample2 %in% single_donor_ASVs$biosample1) %>%
  pull(mixture))


# loop over mixtures
for (mix_id in mixture_ids) {
  ids <- unlist(strsplit(mix_id, "\\+"))
  donor_id <- ids[1]
  recipient_id <- ids[2]
  
  result <- get_colonization(mix_id, donor_id, recipient_id, replicate = 1)
  

  actual_colonizers <- result$colonization_df 
  potential_colonizers <- result$donor_asvs_potential
  
  
  # bind all the mixture rows together
  actual_colonizers_results <- bind_rows(actual_colonizers_results, actual_colonizers)
  potential_colonizers_results <- bind_rows(potential_colonizers_results, potential_colonizers)
}

# loop through donor ids and get potential colonizers. Maybe rewrite a function to extract potential colonizers from the donor ids.


```




**Find differential colonizers**


```{r}

actual_colonizers_results <- actual_colonizers_results %>% 
  mutate(day = str_sub(biosample1, -3), 
         mixture_pair = str_replace_all(mixture, "-\\d{3}", ""))

# We need to group OTUs by their mixture pair, independent of day
diff_colonizers <- actual_colonizers_results %>%
  filter(day %in% c("029", "036")) %>%
  group_by(mixture_pair, OTU) %>%
  summarize(
    colonized_day29 = as.integer(any(day == "029" & actual_colonizer == 1)),
    colonized_day36 = as.integer(any(day == "036" & actual_colonizer == 1)),
    .groups = "drop"
  ) %>%
  mutate(diff_colonizer = as.integer(colonized_day36 & !colonized_day29))


actual_colonizers_results <- actual_colonizers_results %>%
  left_join(diff_colonizers %>% select(mixture_pair, OTU, diff_colonizer),
            by = c("mixture_pair", "OTU")) %>% 
  mutate(diff_colonizer = ifelse(day == "064", 0, diff_colonizer))


```








**Find the strains lost between day 29 and day 36 in the recipient**


```{r}

recipient_ASVs <- recipient_ASVs %>%
  mutate(day = str_extract(biosample1, "\\d{3}"))

recipients_day29 <- recipient_ASVs %>%
  filter(day == "029", replicate == 1) %>%
  select(biosample1, OTU) %>%
  distinct()

recipients_day36 <- recipient_ASVs %>%
  filter(day == "036", replicate == 1) %>%
  select(biosample1, OTU) %>%
  distinct()

lost_strains <- recipients_day29 %>% 
  filter(!(OTU %in% recipients_day36$OTU)) %>% # extract day 29 taxa not in day 36 taxa
  mutate(lost_strain = 1)
  
recipient_ASVs <- recipient_ASVs %>%
  left_join(lost_strains, by = c("biosample1", "OTU")) %>%
  mutate(lost_strain = ifelse(is.na(lost_strain), 0, lost_strain))

recipient_lost <- recipient_ASVs %>% 
  filter(lost_strain == 1)

```





## Are the differential colonizers in the same family as the lost colonizers?


### Create function extract specific mixture and display its composition


```{r}

theme_set(theme_cowplot())

subject_composition <- function(mix_id, donor_id, recipient_id, replicate) {
  
  # Subset mixture and recipient ASVs for the given replicate
  mix_asvs_subset <- mixture_ASVs %>% 
    filter(mixture == mix_id, replicate == replicate)
  
  recipient_asvs_subset <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == replicate)
  
  mixture_ASVs_mix <- actual_colonizers_results %>% 
  filter(mixture == mix_id, replicate == 1) %>% 
  ggplot(aes(x = mixture, y = relAbundance, fill = Family))+
  geom_bar(stat = "identity", color = "black") +
  scale_fill_manual(values = my_colors) +
  labs(
    title = "Mixture",
    x = "",
    y = ""
  ) +
  theme(legend.position = "none",
        axis.line=element_blank(), 
        axis.text.x = element_text(size = 10))


  mixture_ASVs_colonized <- actual_colonizers_results %>% 
    filter(mixture == mix_id, replicate == 1) %>% 
    ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(actual_colonizer))) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
    labs(
      title = "Actual Colonizers",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))
  
    mixture_ASVs_differential <- actual_colonizers_results %>%
    filter(mixture == mix_id, replicate == 1, day == "036") %>%
    ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = factor(diff_colonizer))) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    scale_alpha_manual(values = c("0" = 0, "1" = 1)) +  # colonizer == 0 → transparent
    labs(
      title = "Differential",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(),
          axis.text.x = element_text(size = 10))

  
  lost_ASVs <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == 1, day == "029")%>% 
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family, alpha = factor(lost_strain)))+
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    scale_alpha_manual(values = c("0" = 0, "1" = 1)) +
    labs(
      title = "Lost Strains",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))
  
    recipient_ASVs <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == 1)%>% 
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Recipient",
      x = "",
      y = ""
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))

  donor_ASVs <- single_donor_ASVs %>% 
    filter(biosample1 == donor_id)%>%
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family))+
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Donor",
      x = "",
      y = "Total Relative Abundance"
    ) +
    theme(legend.position = "none",
          axis.line=element_blank(), 
          axis.text.x = element_text(size = 10))


  row_of_plots <- plot_grid(
    donor_ASVs,
    recipient_ASVs,
    mixture_ASVs_mix,
    mixture_ASVs_colonized,
    lost_ASVs,
    mixture_ASVs_differential,
    ncol = 6
  )
  
  # Add a centered title above the row
  final_plot <- plot_grid(
    ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
    row_of_plots,
    ncol = 1,
    rel_heights = c(0.1, 1)  # Reserve top space for title
  )

  return(final_plot)
}


```



**Function call**

```{r}

subject_composition_plot_day29 <- subject_composition("XGB-029+XBA-029", "XGB-029", "XBA-029", 1) + theme_cowplot()
subject_composition_plot_day36 <- subject_composition("XGB-029+XBA-036", "XGB-029", "XBA-036", 1) + theme_cowplot()
# subject_composition_plot_day64 <- subject_composition("XJB-029+XDA-064", "XJB-029", "XDA-064", 1) + theme_cowplot()


combined_plot <- subject_composition_plot_day29 /
                 subject_composition_plot_day36
                 # subject_composition_plot_day64

combined_plot

ggsave(
  filename = file.path(OUTPUTDIR, "compositionsXGBXBA_lost.png"),
  plot = combined_plot,
  width = 20,
  height = 10,
  dpi = 600,
  bg = "white"
)


```












### Bootstrapping



**Extract single-mixture information for potential colonizers**

```{r}

mix_id <- "XGB-029+XBA-029"
mix_pair <- "XGB+XBA"
donor_id <- "XGB-029"
recipient_id <- "XBA-029"

single_result <- get_colonization(mix_id, donor_id, recipient_id, replicate = 1)

single_sample_actual_colonizers <- single_result$colonization_df %>% 
  filter(actual_colonizer == 1)

single_recipient_lost <- recipient_lost %>% 
  filter(biosample1 == recipient_id)

# single_sample_donor_ASVs <- single_donor_ASVs %>% 
#   filter(biosample1 == donor_id)


single_sample_actual_colonizers <- actual_colonizers_results %>% 
  filter(mixture_pair == mix_pair, actual_colonizer == 1, day != "064") %>% 
  distinct(Family, OTU)

donor_communities
```


**Set parameters**
```{r}
n_trials <- 1000
n_sample <- nrow(single_sample_actual_colonizers)  # Number of colonizers observed
```



**Simulate Boostrap Trials**

```{r}

set.seed(123)
trial = 1

bootstrap_results <- map_dfr(1:n_trials, function(trial) {
  boot_sample <- single_sample_actual_colonizers %>%
    slice_sample(n = n_sample, replace = TRUE)
  
  otu_shared <- length(intersect(boot_sample$OTU, single_recipient_lost$OTU)) # get shared taxa between the bootstrap population and the recipient population
  fam_shared <- length(intersect(boot_sample$Family, single_recipient_lost$Family))
  
  tibble(trial = trial, shared_otus = otu_shared, shared_families = fam_shared)
})


length(unique(single_sample_actual_colonizers$OTU))

bootstrap_results

```



**Get observed values**

```{r}

single_sample_diff_colonizers <- actual_colonizers_results %>% 
  filter(mixture_pair == mix_pair, diff_colonizer == 1, day != "064") %>% 
  distinct(Family, OTU)


observed_otus <- length(intersect(single_sample_diff_colonizers$OTU, single_recipient_lost$OTU))
observed_families <- length(intersect(single_sample_diff_colonizers$Family, recipient_lost$Family))

observed_otus
observed_families

intersect(single_sample_diff_colonizers$OTU, recipient_lost$OTU)

bootstrap_results <- bootstrap_results %>%
  mutate(trial = as.character(trial)) %>%
  bind_rows(tibble(
    trial = "Observed",
    shared_otus = observed_otus,
    shared_families = observed_families
  ))



```


**Get p-values**
```{r}
p_otu <- mean(bootstrap_results$shared_otus >= observed_otus)
p_fam <- mean(bootstrap_results$shared_families >= observed_families)

p_otu
p_fam

```




```{r}
ggplot(bootstrap_results, aes(x = shared_otus)) +
  geom_histogram(binwidth = 1, fill = "gray") +
  geom_vline(xintercept = observed_otus, color = "red", linetype = "dashed") +
  labs(title = "Bootstrap OTU Overlap with Lost Taxa",
       subtitle = paste("Observed =", observed_otus, ", p =", signif(p_otu, 3)))


ggplot(bootstrap_results, aes(x = shared_families)) +
  geom_histogram(binwidth = 1, fill = "gray") +
  geom_vline(xintercept = observed_families, color = "red", linetype = "dashed") +
  labs(title = "Bootstrap Families Overlap with Lost Taxa",
       subtitle = paste("Observed =", observed_families, ", p =", signif(p_fam, 3)))
```

```{r}
recipient_lost %>% 
  group_by(biosample1) %>% 
  summarize(total_lost = sum(lost_strain))
```




