---
title: "e0029 Divergence and Colonization Analysis"
output: html_document
date: "2025-05-21"
---


# Load in packages

```{r, warning=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(readr)
library(ggrepel)
library(purrr)
library(patchwork)
library(foreach)
library(forcats)
library(cowplot)
```


# Load in the Household Project dataset

```{r}

household_data <- read.table("C:/Users/anaes/OneDrive/UCI_Spring25/rotation/e0026-e0029-e0030.txt", header = TRUE)

# Divide dataset into separate tables by experiment
e0029 <- filter(household_data, str_detect(household_data$sample, "e0029"))

e0026 <- filter(household_data, str_detect(household_data$sample, "e0026"))

# Extract subject, day, household information
e0029_clean <- e0029 %>%
  mutate(
    subject = str_sub(biosample1, 1, -5),
    day = str_sub(biosample1, -3),
    household = str_sub(biosample1, 1, -6)
  )

# Specify output directory

OUTPUTDIR <- "./e0029_outputs"

# Define color palette

my_colors <- readRDS("C:/Spring-Rotation-2025/scratch/familyColorPalette.rds") 

```



### Get beta diversity


**Create phyloseq object**


```{r}

data <- readRDS("../../data/ps_all.rds")

hp_data_clean <- prune_taxa(taxa_sums(data) > 0, data)

e0026_e0029_obj <- subset_samples(
  hp_data_clean,
  experiment == "e0029" | 
    experiment == "e0026"
)

# e0026_e0029_obj <- subset_samples(e0026_e0029_obj,
#     biosample1 != "blank" &
#     biosample1 != "B-mix" &
#     !grepl("\\+", biosample2))

```


```{r}
# # List the distance-calculation methods to be used.
# distMethods <- c("jsd","bray","jaccard")
# 
# # Write a function to calculate a distance matrix using the specified method
# # and convert the data into tidy format.
# calculateBeta <- function(data, distMethod) {
#   # Calculate the distance matrix using the specified method.
#   betaRaw <- distance(data, method=distMethod)
# 
#   # Convert distance matrix to a dataframe.
#   beta <- as.matrix(betaRaw)
#   beta <- as.data.frame(beta)
#   beta$sample1 <- rownames(beta)
#   # Tidy the dataframe.
#   beta <- beta %>%
# 	pivot_longer(-sample1, names_to="sample2", values_to="value")
#   beta <- beta %>%
# 	filter(sample1 != sample2) %>%
# 	mutate(method=distMethod)
# }
# 
# # Calculate the distance matrix for all of the specified methods on the species abundances.
# # Combine the distance matrices for all methods.
# betaSpecies <- do.call(rbind, lapply(distMethods, function(distMethod) {
#   print(distMethod)
#   calculateBeta(e0026_e0029_obj, distMethod)
# }))
# # Export the distance matrix generated for all of the sample pairs
# # using all of the specified methods.
# write_delim(betaSpecies, paste0("e0026_e0029_speciesBeta.txt.gz"))


```



### JSD between all possible combinations



```{r, warning=FALSE}

beta_diversity <- read_delim("./e0026_e0029_speciesBeta.txt.gz")
data(beta_diversity)


beta_diversity_wide <- beta_diversity %>%
  select(sample1, sample2, method, value) %>%
  pivot_wider(names_from = method, values_from = value)

beta_diversity_wide <- beta_diversity_wide %>%
  filter(!grepl("e0029", sample1), !grepl("e0029", sample2))

```




```{r}

recipients <- c("XBA-029", "XBA-036", "XBA-064", "XDA-029", "XDA-036", "XDA-064", "XEA-029", "XEA-036", "XEA-064", "XKA-029", "XKA-036", "XKA-064")

e0026_filtered <- e0026 %>% 
  select(sample, biosample1) %>% 
  unique()

donor_communities <- e0029 %>%
  filter(biosample2 != "blank", 
         biosample2 != "B-mix",
         !grepl("\\+", biosample2)) %>%
  pull(biosample2) %>%
  unique()

# join sample1 metadata and check
e0026_e0029_beta_1 <- beta_diversity_wide %>%
  left_join(e0026_filtered, by = c("sample1" = "sample")) %>% 
  filter(biosample1 %in% recipients)

# if the columns exist, rename them
e0026_e0029_beta_1 <- e0026_e0029_beta_1 %>%
    dplyr::rename(
      component1 = biosample1
    )

# Join sample2 metadata
e0026_e0029_beta_final <- e0026_e0029_beta_1 %>%
  left_join(e0026_filtered, by = c("sample2" = "sample")) %>% 
  filter(biosample1 %in% donor_communities)


# Rename columns to have biosample1 and biosample
e0026_e0029_beta_final <- e0026_e0029_beta_final %>%
    dplyr::rename(
      biosample1 = component1,
      biosample2 = biosample1
    )


```


**Define a function to extract passage information from raw sample ID**

```{r}

extract_passage <- function(x) {
  ifelse(
    str_detect(x, "-mix-"),
    NA_character_,
    str_extract(x, "-[A-Z]-[08]-[A-Z0-9]+") %>%
      str_extract("[08]")
  )
}

# Add passage extraction
e0026_e0029_beta_final <- e0026_e0029_beta_final %>%
  mutate(
    passage1 = extract_passage(sample1),
    passage2 = extract_passage(sample2)
  ) %>% 
  filter(passage1 == 8, passage1 == passage2)

e0026_e0029_beta_final <- e0026_e0029_beta_final %>% 
  select(-passage1, -passage2, -bray, -jaccard) %>% 
  mutate(mixture = paste(biosample2, biosample1, sep = "+"))
  
```


**Plot distances in a heatmap**

```{r}

theme_set(theme_cowplot())

beta_diversity_mixture_plot <- ggplot(e0026_e0029_beta_final, aes(x = biosample2, y = biosample1, fill = jsd)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  labs(x = "Donor", y = "Recipient", fill = "JSD",
       title = "JSD Heatmap Across Donor–Recipient Pairs")

ggsave(
  filename = file.path(OUTPUTDIR, "diversity_heatmap_plot.png"),
  plot = beta_diversity_mixture_plot,
  width = 10,
  height = 10,
  dpi = 600,
  bg = "white"
)

beta_diversity_mixture_plot


```


**Boxplot of JSDs faceted by day**

```{r}

theme_set(theme_cowplot())

e0026_e0029_beta_final <- e0026_e0029_beta_final %>% 
  mutate(day = str_sub(biosample1, -3), subject = str_sub(biosample1, 1, -5))

diversity_boxplot <- e0026_e0029_beta_final %>% 
  ggplot(aes(x = subject, y = jsd, fill = subject)) + 
  geom_boxplot()+
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2), size = 1, alpha = 0.6) +
  labs(
    title = "Divergence between donor-recipient pair distributions by day",
    x = "",
    y = "JSD",
    fill = ""
  )+
  theme(
    axis.text.x = element_text(angle = 45, vjust = 0.5, hjust = 1, size = 8)
  )+
  facet_wrap(~day)


ggsave(
  filename = file.path(OUTPUTDIR, "diversity_boxplot_mixtures.png"),
  plot = diversity_boxplot,
  width = 10,
  height = 10,
  dpi = 600,
  bg = "white"
)



```



**Computing colonization as an efficacy metric**

```{r}

get_colonization <- function(mix_id, donor_id, recipient_id, replicate) {
  mix_asvs_subset <- mixture_ASVs %>% 
    filter(mixture == mix_id, replicate == replicate)
  
  recipient_asvs_subset <- recipient_ASVs %>% 
    filter(biosample1 == recipient_id, replicate == replicate)
  
  donor_asvs_subset <- single_donor_ASVs %>%  
    filter(biosample1 == donor_id)
  
  # potential colonizers are those present in the donor but not in the recipient
  potential_colonizers <- donor_asvs_subset$OTU[!(donor_asvs_subset$OTU %in% recipient_asvs_subset$OTU)]

  # actual colonizers are the subset of potentials that did make it into the mix
  mix_asvs_colonization <- mix_asvs_subset %>% 
    mutate(actual_colonizer = ifelse(OTU %in% potential_colonizers, 1, 0))
  
  # return both the dataframe and potential colonizer list
  return(list(
    colonization_df = mix_asvs_colonization,
    potential_colonizers = potential_colonizers
  ))
}


```




**Loop through all mixtures to get final e0029 colonization dataset**

```{r}

# initialize results table
colonization_prop_results <- tibble()

# get list of mixture IDs
mixture_ids <- unique(mixture_ASVs %>%
  filter(biosample2 %in% single_donor_ASVs$biosample1) %>%
  pull(mixture))

# loop over mixtures
for (mix_id in mixture_ids) {
  ids <- unlist(strsplit(mix_id, "\\+"))
  donor_id <- ids[1]
  recipient_id <- ids[2]
  
  result <- get_colonization(mix_id, donor_id, recipient_id, replicate = 1)
  
  # count actual and potential colonizers
  n_potential_colonizers <- length(result$potential_colonizers)
  n_actual_colonizers <- result$colonization_df %>%
  filter(actual_colonizer == 1) %>%
  distinct(OTU) %>%
  nrow()
  
  # store all values in a one-row tibble
  prop_row <- tibble(
    mixture = mix_id,
    biosample2 = donor_id,
    biosample1 = recipient_id,
    n_potential_colonizers = n_potential_colonizers,
    n_actual_colonizers = n_actual_colonizers,
    prop_colonizers = n_actual_colonizers / n_potential_colonizers
  )
  
  # bind all the mixture rows together
  colonization_prop_results <- bind_rows(colonization_prop_results, prop_row)
}


```


**Visualize colonization**

```{r}
theme_set(theme_cowplot())

prop_colonization_plot <- ggplot(colonization_prop_results, aes(x = donor_id, y = recipient_id, fill = prop_colonizers)) +
  geom_tile() +
  scale_fill_viridis_c(option = "A") +
  labs(x = "Donor", y = "Recipient", fill = "Colonization Efficacy",
       title = "Colonization Efficacy Across Mixtures")

ggsave(
  filename = file.path(OUTPUTDIR, "prop_colonization_plot.png"),
  plot = prop_colonization_plot,
  width = 12,
  height = 10,
  dpi = 600,
  bg = "white"
)

```



**Plot Colonization against JSD**

```{r}

# join in the two datasets

jsd_data <- e0026_e0029_beta_final %>% 
  select(mixture, jsd, day)

jsd_colonization <- colonization_prop_results %>%
  left_join(jsd_data, by = c("mixture" = "mixture"))

jsd_colonization_plot <- jsd_colonization %>% 
  ggplot(aes(x = jsd, y = prop_colonizers, color = factor(day)))+
  geom_point()+
  labs(
    x = "Donor-Recipient JSD",
    y = "Colonization Efficacy",
    title = "Colonization Efficacy vs. Donor-Recipient Divergence"
  )

ggsave(
  filename = file.path(OUTPUTDIR, "jsd_colonization_plot.png"),
  plot = jsd_colonization_plot,
  width = 10,
  height = 8,
  dpi = 600,
  bg = "white"
)
  
  

```



### Differential Colonization


**Find differential colonizers**

```{r}



```





























**Composition Plots**

```{r}
theme_set(theme_cowplot())

subject_composition <- function(mix_id, donor_id, recipient_id, replicate) {
  
  # Run colonization analysis for this donor-recipient pair
  result <- get_colonization(mix_id, donor_id, recipient_id, replicate)
  colonization_df <- result$colonization_df
  
  # Mixture composition
  mixture_ASVs_mix <- colonization_df %>%
    ggplot(aes(x = mixture, y = relAbundance, fill = Family)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Mixture",
      x = "",
      y = ""
    ) +
    theme(
      legend.position = "none",
      axis.line = element_blank(),
      axis.text.x = element_text(size = 10)
    )

  # Colonized subset (highlighted)
  mixture_ASVs_colonized <- colonization_df %>%
    mutate(colonization = factor(actual_colonizer)) %>%
    ggplot(aes(x = mixture, y = relAbundance, fill = Family, alpha = colonization)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    scale_alpha_manual(values = c("0" = 0, "1" = 1)) +
    labs(
      title = "Colonized",
      x = "",
      y = ""
    ) +
    theme(
      legend.position = "none",
      axis.line = element_blank(),
      axis.text.x = element_text(size = 10)
    )

  # Recipient composition
  recipient_plot <- recipient_ASVs %>%
    filter(biosample1 == recipient_id, replicate == replicate) %>%
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Recipient",
      x = "",
      y = ""
    ) +
    theme(
      legend.position = "none",
      axis.line = element_blank(),
      axis.text.x = element_text(size = 10)
    )

  # Donor composition
  donor_plot <- single_donor_ASVs %>%
    filter(biosample1 == donor_id) %>%
    ggplot(aes(x = biosample1, y = relAbundance, fill = Family)) +
    geom_bar(stat = "identity", color = "black") +
    scale_fill_manual(values = my_colors) +
    labs(
      title = "Donor",
      x = "",
      y = "Total Relative Abundance"
    ) +
    theme(
      legend.position = "none",
      axis.line = element_blank(),
      axis.text.x = element_text(size = 10)
    )

  # Assemble panel
  row_of_plots <- plot_grid(
    donor_plot,
    recipient_plot,
    mixture_ASVs_mix,
    mixture_ASVs_colonized,
    ncol = 4
  )

  final_plot <- plot_grid(
    ggdraw() + draw_label("", fontface = "bold", size = 12, hjust = 0.5),
    row_of_plots,
    ncol = 1,
    rel_heights = c(0.1, 1)
  )

  return(final_plot)
}

```



**Function call**

```{r}

subject_composition_plot_day29 <- subject_composition("XJB-029+XEA-029", "XJB-029", "XEA-029", 1) + theme_cowplot()
subject_composition_plot_day36 <- subject_composition("XJB-029+XEA-036", "XJB-029", "XEA-036", 1) + theme_cowplot()
subject_composition_plot_day64 <- subject_composition("XJB-029+XEA-064", "XJB-029", "XEA-064", 1) + theme_cowplot()

combined_plot <- subject_composition_plot_day29 /
                 subject_composition_plot_day36 /
                 subject_composition_plot_day64

ggsave(
  filename = file.path(OUTPUTDIR, "compositions_XKBXEA.png"),
  plot = combined_plot,
  width = 10,
  height = 10,
  dpi = 600,
  bg = "white"
)


```


